Supplementary Information

### 2021-01-29 doc/manuscript/MEE_manuscript.Rmd
work on sampling and survivorship biases
#109

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig. `r year_fig`).
This finding was consistent with an independent analysis of growth-$DBH$ trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), and results were generally consistent regardless of the growth metric analyzed (Figs. `r all_GLS`), indicating that our approach is robust in its analytical structure.
Our results are, however, still subject to the influence of some potential sampling and survivorship biases [Fig. `r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014], evaluation of which is beyond the scope of the current analysis.
At a minimum, our approach provides a more appropriate alternative to two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
Indeed, in a comparative analysis of methods to detect growth trends, approaches similar to that employed here (regional curve standardization) performed best at growth trend detection [@peters_detecting_2015], supporting the approach used here as current best practice *[but check @hember_tree_2019].*
Yet, the trends detected here remain potentially subject to a variety of sampling and survivorship biases that can influence overall trends in positive or negative directions (*REFS*) [Fig. `r year_fig`; @brienen_detecting_2012; @brienen_tree_2017].

...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].

@@

Signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
The majority of negative growth trends observed here, *to the extent that they are robust to potential sampling and survivorship biases*, are probably attributable to increasing competition as secondary stands age. 
Several of our sites (**HKK, SCBI, LDW, HF, ...** *Coauthors, I need help here.*) are known to have suffered major disturbances within or shortly prior to the analysis time frame [Appendix `r site_details_appendix`; REFS: @baker_disturbance_2005; @bourg_initial_2013], and growth declines at these sites are consistent with increasing competition as young stands mature.
*Indeed, stand succession has been recognized as a driver of growth declines in tree ring records... (REFS).*
Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.

>>>>

Growth rates changed directionally for the majority (`r percent_sig_neg_trend_of_analyzed`, depending on growth metric) of site-species combinations selected for analysis based on the sampling representation of $DBH$ by $year$ (Fig. `r year_fig`).
These results were similar across growth metrics (Figs. `r all_GLS`), and consistent with an independent analysis of growth-$DBH$ trends by decade (Figs. `r year_fig`, `r decadal_analysis_figs_all`), indicating that our approach is robust in its analytical structure.
Indeed, in a comparative analysis of several methods commonly used to detect growth trends, approaches similar to that employed here (regional curve standardization) performed better at growth trend detection [@peters_detecting_2015] than two of the most commonly used methods for analyzing growth trends, conservative detrending and basal area correction [see @peters_detecting_2015], both of which would fail to capture the observed trends in $RW$ and $BAI$ with $DBH$ (Fig. `r size`).
For instance, based on our finding that $BAI$ tends to rise and fall with $DBH$, basal area correction would generally be biased towards finding positive trends for smaller trees and negative trends for larger trees.
However, are results remain subject to the influence of some potential sampling and survivorship biases that can influence overall trends in positive or negative directions [Fig. `r year_fig`; @brienen_detecting_2012; @brienen_tree_2017; @nehrbassahles_influence_2014].
While we have made efforts to control for the most severe potential biases (see Methods), and while the statistical model applied here in theory accounts for persistent growth differences among trees through a random effect of tree, some potential biases are difficult, if not impossible, to control for [@brienen_tree_2017].
The observed trends must therefore be interpreted with caution, and in light of the fact that growth trends inferred from tree rings can be strongly shaped by tree demography [@brienen_tree_2017].

*...  corroborates previous analyses from HKK showing growth decreases over time [@nock_longterm_2011; @van_der_sleen_no_2015; @groenendijk_no_2015, but see @brienen_tree_2017].*

@@

To the extent that sampling and survivorship biases are minimal, signals of changing growth rate over time are attributable to some combination of stand dynamics (e.g., recruitment and succession, changing stand structure) and environmental drivers (e.g., climate drivers other than those selected by *climwin*; atmospheric CO~2~; deposition of sulfur dioxide, SO~2~, and nitrogen oxides, NO~x~).
The majority of negative growth trends observed here are probably attributable to successional stand dynamics as cohorts and stands age. 
For species exhibiting a pulse of recruitment in the past and little subsequent recruitment (e.g., **SPECIES, FIGURES**), variation in growth rates among individuals would naturally lead to a trend of declining growth, as faster-growing individuals reach various size thresholds earlier [@brienen_tree_2017].
Such a trend is not indicative of a directional response to changing environmental conditions that would be generalizable across stands (e.g., rising atmospheric CO~2~), but does indicate a decline of the species within the stand. 
This can occur, for example, when a light-demanding species is no longer recruiting under a closed canopy, which is the likely explanation for declines in several species (e.g., **SPECIES, FIGURES/TABLES**).
In secondary stands, particularly those where the majority of trees established at a similar time (**SCBI, HF,...**; Appendix `r site_details_appendix` **FIGURES**), growth declines are consistent with the tendency for faster tree growth during early succession (**REFS**), and with increasing competition as young stands mature (**REFS**).
Even within mature forests, light-demanding species that establish in gaps (e.g., *Jacaranda copaia* at BCNM, *Populus tremuloides* at CB; Table `r species_table`) would tend to experience an increasingly competitive environment.
For more shade-tolerant species in stands with no known disturbance within the past 1.5 centuries (**ZOF?, NM?, CB, SC**), trends would represent more gradual successional changes or responses to global change.
At Cedar Breaks, which has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data), mixed growth trends probably reflect later successional changes in a mixed-disturbance forest and/or shifting competitive advantages, perhaps in part driven by changing environmental conditions.
